Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-16 paper/paper.tex
write comparison and conclusion (with fixmes) Tested everything on bennet.

\section{Model, Methods, and Cell Shapes}
We implement the reaction-diffusion model of Huang \emph{et
  al.}~\cite{huang2003dynamic}.  Figure~\ref{fig:reactions} shows the
reaction process.  The cytoplasmic MinD:ADP complex undergoes
nucleotide exchange and is changed into the MinD:ATP complex.  This
will naturally diffuse and attach to the cell membrane.  A cytoplasmic
MinE will attach to the wall bound MinD:ATP complex and after a time
will activate ATP hydrolosis.  This breaks up the complex, releasing
MinE, phosphate, and MinD:ADP back into the cytoplasm.  The MinD:ADP
will undergo nucleotide exchange and begin again the cyclic process.
The model is defined by a set of five reaction-diffusion equations:

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes and sizes.  Here we present a selection of
these, first naturally occuring pill-shaped cells and then a number of
flattened out shapes which reflect the experiments of Mannik \emph{et
  al.}, in which bacteria are confined within a thin slit of height
$0.25\micron$ ~\cite{mannik2012robustness}. Viewed from the top down
the cells will show the shapes described below and viewed from the
side they show at their edges a semicircular protrusion (one may
imagine the shape of a pancake). We have simulated and analyzed a
number of these flattened shapes in addition to those shown below,
including equilateral and iscosolese triangle shapes, shapes shaped
like greater than signs, '$>$', and four-pronged 'starfish' shapes.
For all of these we have varied the cell size, expanding the two
dimensions that are tangent to the flattened surface from very small,
'steady state' scales (which we discuss below) to scales twice that of
the published Mannik shapes.  Our conclusions are based on all of this
data, but we present here specific, illustrative examples: two
flattened shapes that are designed to replicate those published by
Mannik and two 'stadium' shapes that respectively have the same
quadrupole moments of the two Mannik shapes.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{../paper/plot-ave}
  \caption{Arrows depicting global maxima in space and local
    maxima in
    time overlayed on color plots of time averaged total MinD density.
    A density threshold is chosen and only maxima above this threshold
    are shown. The plots serve to illistrate what is observed in our
    movies.  The simulation time covered in these is 1000 seconds
    \fixme{this is not true yet - need to change the data once its
      finished right now 2 of them are 1000 the others are like 980}.
    The top row shows plots published by Mannik of the MinD maxima
    behavoir and the bottom two rows show our pancake shapes that are
    designed to have the same shape.  The middle row shows results
    from the stochastic simulation and the bottom shows results from
    the deterministic.  We can see that the stochastic method shows
    behavoir that is similar to experiment while the deterministic
    method shows maximization that is too regular and robust.}
  \label{randst-plot-ave}
\end{figure*}

@@

We have made and analyzed movies out of the MinD
concentration that are constructed as those in Figure~\ref{image-p},
integrated along the flattened axis, in which each frame is seperated
in simulation time by a half second. Figure~\ref{randst-plot-ave}
illustrates what is seen in these movies. In his work Mannik presents
arrow plots in which the arrow heads show the location of sequential
MinD maxima within the cell.  These experimental plots for two abnormal cell
shapes are shown in the upper `experiment' section of
Figure~\ref{randst-plot-ave}.  Following this, we have created plots
that show maxima that are global in space and local in time.  Of these
maxima we only show those which have greater concentration than a
chosen cutoff concentration, and also ignore maxima that appear
temporaly and spatially adjacent to preceding maxima, in order to
stop single spikes from appearing as many adjecent ones.  The overall
effect is to accurately describe what is seen in our movies.

We have found that the deterministic simulations of the protein
reaction sequence discussed above results in robust oscillatory
behavoir of polar selection and oscillation in not only symmetrical
shapes, such as the wildtype pill shape, but also in very
assymetrical, flatttened cells, such as those studied by Mannik.
After the lower size stability limit of
section~\ref{sec:stability-analysis} \fixme{this needs to be covered}
has been surpassed, this qualitative behavoir doesn't seem to have
size dependence until cells reach sizes that have more than twice the
dimensions of the Mannik cells.  At this point the robust
oscillations, while still evident, begin to lose their consistency.
The Mannik reported cells already seem to be rather large (wild type
cells are typically smaller than those observed my Mannik
\cite{kubitschek1990cell,kubitschek1968linear,mannik2012robustness}),
so we left off simulating cells any larger than twice this size.

Looking at Mannik's shape A in Figure~\ref{randst-plot-ave}, we can
see that the deterministic simulation selects out two poles, one at
either end of the cell, and that the MinD does not ever maximize in
any other location.  All of the flattened cells shown in the figure
span a time of 350s (as do Mannik's plots), which corresponds to
\fixme{number of periods} periods for shape A.  The movies show sharp
peaks at these poles, followed by a spreading out of the MinD out into
the cell, with no real bias towards any location, until it aggregates
again at the opposite pole, maximizing there.  The pattern is repeated
again and again with surprising consistency, \fixme{talk about the
  time correlation for the deterministic, once the plots are added.}

The results of these deterministic simulations for shape A differ
significantly with Mannik's experimental results that show much less
regular behavoir, with a large variation in the location of maxima,
shown in the experiment section of Figure~\ref{randst-plot-ave}.

The deterministic results of shape B are slightly different, as can be
seen in the figure.  Here we can see that the proteins maximize on the
rightward cell wall while on the way to their corner maxima locations.
Watching the movies, the proteins seem to travel along the wall, from
corner to corner and back.  This behavoir is still more irregular than
that of the Mannik experiments, which seem to show some right-left
back and forth behavoir in a few areas, as opposed to the straight
pole to pole (traveling along the wall).

These comparisons illustrate our first finding, which is that in
irregular cell shapes, the deterministic method of simulation is indeed
inadequate in representing and predicting experimental behavoir.

On the other hand, our stochastic simulations show both robust pole to
pole oscillations in the symmetric cells and irregular behavoir in the
assymetrical shapes, agreeing well with Mannik's experimental
findings.

Figure~\ref{arrow-pill-stadium} and Figure~\ref{randst-plot-ave} show
color plots of time-averaged MinD overlayed by arrows depicting
sequential maxima that are global in space and local in time.  Our
stochastic method movies (and arrow plots of
Figure~\ref{randst-plot-ave}) show irregular location of maxima that
is similar to what is observed experimentally by Mannik. There is a
general bias towards back and forth behavoir, which can be seen in the
box plots of Figure~\ref{box-mannik}, but the behavoir is irregular
and the location of the maxima vary a significant amount.  On the
other hand, the deterministic simulations show behavoir that is much
too regular when compared with experiment, exhibiting locations of
maxima that do not vary.  The deterministic method is seen to be
inadequate.

Judging that at this point that we have a fair idea of how our two
methods compare with experiment, we investigate the limits of the
break down of regular activity, by simulating large, regular,
'stadium' shapes.  These shapes have the flattened, pancake like form
of the assymetric cells, but from the top down they look as pills
do. In two dimensions they are composed of a rectangle with two half
ellipse connected on either end.  They are also made to have a large
volume that is similar to the Mannik shapes.

We find that both the stochastic and the deterministic solutions show
regular back and forth oscillations in the stadium shaped cells, as
can be seen Figure~\ref{box-stadium}.  This suggests that it is the
irregular, assymetric shape of the Mannik cells, not their size or
their flattened nature, that causes these cells to exhibit
irregularity in MinD oscillations.

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-5_50-1_00-0_00-15_00-full_array/plots/box-plot_D}
%%   \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-5_50-1_00-0_00-15_00-exact/plots/box-plot_D}
%%   \caption{A comparison of our box plots that show the similarity
%%     between the stadium shapes simulated stochastically and
%%     deterministically.  Both show regular oscillatory maxima with
%%     similar periods.  This is in contrast to our boxpplot comparison
%%     of one of the Mannik shapes in Figure~\ref{box-mannik}}
%%   \label{box-stadium}
%% \end{figure}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/box-plot_D}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-exact/plots/box-plot_D}
%%   \caption{A comparison of our box plots that show a disparity between
%%     the behavoir of the stochastic and deterministic simulations in
%%     our Mannik cells.  The deterministic show a regular, robust
%%     oscillatory behavoir between the two ends of the cell. The
%%     stochastic simulation do show a general back and forth tendency
%%     between two ends of the cell, but it is not robust and tends to
%%     break down.}
%%   \label{box-mannik}
%% \end{figure}
%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-full_array/plots/box-plot_D}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-exact/plots/box-plot_D}
%%   \caption{A comparison of our box plots that show a disparity between
%%     the behavoir of the stochastic and deterministic simulations in
%%     our Mannik cells.  The deterministic show a regular, robust
%%     oscillatory behavoir between the two ends of the cell. The
%%     stochastic simulation do show a general back and forth tendency
%%     between two ends of the cell, but it is not robust and tends to
%%     break down.}
%%   \label{box-mannik}
%% \end{figure}

\fixme{specifically design other arrow plot (with many comparisons) to
  illustrate point}

Figure~\ref{arrow-compare-mannik} shows plots of total MinD
concentration maxima that are global in space and local in time.  The
tip of an arrow represents one of these maxima and the adjacent arrow
(whose tail is touching the previous arrow's tip) points to the next
such maxima in time.  Mannik \emph{et al.} have published similar
plots of two of their cells, and we have attempted to replicate the
shape of these cells in our simulation in order to compare.  We find
that when we replicate the size as well as shape of their cells as
near as possible, we do not see the disordered protein behavoir that
they observe.  Instead we see a robust ability of the cell to find two
prefered pole locations and to oscillate maxima back and forth very
steadily between them.  When we increase the size of our cells, we
indeed see multiple positions of maxima develop, but we never observe
the random behavoir that they have observed experimentally.  This is
perhaps due to the exact, continuous method we use when conducting
simulations.  Stochastic methods would perhaps show more random
behavoir at these sizes.  We consider this to be a limitation of the
exact solution method.

\fixme{should I say anything about different/large sizes?}

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation-rl.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation-rl.pdf}
  \caption{Correlation for A shapes}
  \label{corr-A}
\end{figure}

\section{Conclusion}
No conclusion :(
\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-full_array/plots/correlation-rl.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_92-1_18-0_00-15_00-full_array/plots/correlation-rl.pdf}
  \caption{Correlation for B shapes}
  \label{corr-B}
\end{figure}

@@

>>>>

\section{Model, Methods, and Cell Shapes}\label{sec:model-method-shapes}
We implement the reaction-diffusion model of Huang \emph{et
  al.}~\cite{huang2003dynamic}.  Figure~\ref{fig:reactions} shows the
reaction process.  The cytoplasmic MinD:ADP complex undergoes
nucleotide exchange and is changed into the MinD:ATP complex.  This
will naturally diffuse and attach to the cell membrane.  A cytoplasmic
MinE will attach to the wall bound MinD:ATP complex and after a time
will activate ATP hydrolosis.  This breaks up the complex, releasing
MinE, phosphate, and MinD:ADP back into the cytoplasm.  The MinD:ADP
will undergo nucleotide exchange and begin again the cyclic process.
The model is defined by a set of five reaction-diffusion equations:

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes and sizes.  Here we present a selection of
these, first naturally occuring pill-shaped cells and then a number of
flattened out shapes which reflect the experiments of Mannik \emph{et
  al.}, in which bacteria are confined within a thin slit of height
$0.25\micron$ ~\cite{mannik2012robustness}. Viewed from the top down
the cells will show the shapes described below and viewed from the
side they show at their edges a semicircular protrusion (one may
imagine the shape of a pancake). We have simulated and analyzed a
number of these flattened shapes in addition to those shown below,
including equilateral and iscosolese triangle shapes, shapes shaped
like greater than signs, '$>$', and four-pronged 'starfish' shapes.
For all of these we have varied the cell size, expanding the two
dimensions that are tangent to the flattened surface from very small,
'steady state' scales (which we discuss below) to scales twice that of
the published Mannik shapes.  Our conclusions are based on all of this
data, but we present here specific, illustrative examples: two
flattened shapes that are designed to replicate those published by
Mannik and two `stadium' shapes that respectively have the same
quadrupole moments of the two Mannik shapes.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{../paper/plot-ave}
  \caption{We display here arrows depicting successive maxima in space
    and and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is \fixme{340} seconds, which is the same
    period of time depicted in the experimental data plots of Mannik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild type pill
    shape, we only cover \fixme{???} seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by Mannik of the MinD maxima behavoir
    and the bottom two rows show our simulations using the stochastic
    and deterministic algorithms, respectively.  We simulated
    approximations to the two shapes observed by Mannik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

As describe in Section~\ref{sec:model-method-shapes}, we will focus on
simulation of four flattened cell shapes: two shapes copied from
Mannik's experiment~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}).  These stadium shapes enable us to
distinguish between the effect of cellular irregularity and that of
overall size and aspect ratio.  We note that shape \emph{B} is more
irregular than shape \emph{A}, and in particular features a prominant
region of concavity on its left-hand side.  For each of these shapes
(in addition to the wild type pill shape discussed previously) we have
simulated in excess of \fixme{???} seconds of evolution of the MinD
system using both the deterministic model of Huang~\fixme{cite} and
the stochastic model of Huang~\fixme{cite}.

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the results of Mannik exemplified in
arrow plots in which the arrow heads show the location of sequential
MinD maxima within the cell (over the same time
interval)~\cite{mannik2012robustness}.  This comparison is shown in
Fig.~\ref{randst-plot-ave}.  In addition to arrows between successive
maxima in space and time, we plot as a colored background the density
of MinD protiens averaged over the same time period.  We note that we
have manually verified that our (computer-generated) arrow plots also
reflect a human interpretation of a movie of the same data.  Finally,
for comparison we present the same plot for a wild-type cell, with a
time period of \fixme{???}  seconds to account for its short period of
oscillation.  In every case, including the wild type pill shape, we
see irregularity in the location of the maxima when using the
stochastic approach.  The deterministic method shows uniformly bipolar
oscillation, with the exception of shape \emph{B}, which develops weak
intermediate maxima on the right-hand edge of the cell.

%% We found that the deterministic simulations of the MinD system results
%% in robust and regular oscillatory behavoir of polar selection and
%% oscillation in not only symmetrical shapes, such as the wildtype pill
%% shape, but also in very assymetrical, flatttened cells, such as those
%% observed by Mannik.  We also examined larger and smaller versions of
%% these shapes, and found this behavior to be very robust, from the
%% minimum size to sustain oscillation (around \fixme{???} $\mu$m) up to
%% around twice the cell size reported by Mannik.  At larger sizes than
%% this, less regular oscillations occur using the deterministic method,
%% but we note that the cells reported by Mannik are already considerably
%% larger in volume than typical wild type
%% cells~\cite{kubitschek1990cell,kubitschek1968linear,mannik2012robustness}.

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation-rl.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation-rl.pdf}
  \caption{Correlation for A shapes}
  \label{fig:corr-pancake-A}
\end{figure}

The predictions of the stochastic as seen in
Fig.~\ref{randst-plot-ave} are roughly similar to the experimental
results in terms of irregularity of the locations of maxima, although
there are locations on the edges of the cells where experiment shows
maxima occuring that we never see in our simulations, suggesting that
the model does not precisely reflect the experiment.  We also note
that the stadium shapes in Fig.~\ref{randst-plot-ave} appear
qualitatively similar to the irregular shapes in terms of the
locations of their maxima.  From these results, we conclude that the
deterministic method is inadequate to explain experimental
observations of the locations of density maxima of the MinD protein.

However, Fig.~\ref{randst-plot-ave} leaves unclear the importance of
irregularity versus overall thickness and aspect ratio in developing
an understanding of the origin of the experimentally observed
behavior.  Although in the movies the stadiums appear to display a
more regular bipolar oscillation, the arrow plots do not provide a
convincing demonstration.  To illustrate the difference in behavior
between shape \emph{A} and stadium \emph{A}, we return in
Fig.~\ref{fig:corr-pancake-A} to the correlation function between the
number of proteins in a segment at the top of the cell and the number
of proteins in a segment at the bottom of the cell.  We computed these
correlation functions using over \fixme{???} seconds of simulation.
We found with the stochastic method that the experimental shape
\emph{A} has a coherence time of under 4 periods, while the
corresponding stadium \emph{A} has a coherence time of 10 periods.  In
each case, the deterministic method is perfectly periodic. This
confirms that using the stochastic method the stadium shape does
indeed have a far more regular bipolar oscillation than the irregular
shape \emph{A}.  When we compare shape \emph{B} and stadium \emph{B},
we see a similar effect, although both coherence times are somewhat
smaller (2.7 and 6.2, respectively).  From these results we conclude
that the irregularity seen experimentally is indeed exacerbated by the
irregular shapes taken on by the cells forced into a thin slit.

\section{Conclusion}
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}.  We find that the
deterministic method~\fixme{cite} predicts behavior that completely
contradicts what is seen experimentally: it predicts stable and
periodic bipolar oscillation, with a minor intermediate maximum
appearing in shape \emph{B}.  Thus we conclude that this method is
inconsistent with experiment, and should not be relied upon for
predictions of the behavior of the MinD system.

In contrast, the stochastic method~\fixme{cite} predicts spatially
irregular formation of maxima for all the flattened cells that we
studied.  In order to better understand the irregular oscillations
observed in experiment, we simulated flattened ``stadium'' shapes in
order to distinguish between the effect of flattening and the effect
of irregular shape.  We found that although the stadium cells
exhibited somewhat more irregular locations of maxima than the
wild-type cells, their oscillation was far more periodic and bipolar
than the irregular cell shapes observed experimentally.  Thus we
confirm that the disruption of the MinD oscillations seen by Mannik
\emph{et al.} is largely due to the irregularity of cell shapes
observed.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-full_array/plots/correlation-rl.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_92-1_18-0_00-15_00-full_array/plots/correlation-rl.pdf}
  \caption{Correlation for B shapes}
  \label{corr-B}
\end{figure}
